Fast algebraic filtering of surfaces from 3D medical images with Julia

### 2019-06-11 src/background.tex
posted contents for implementation section

A \emph{representation scheme} for solid modeling is a mapping between a space of mathematical models and a space of symbolic representations, often generated by a formal grammar.
Solid pointsets (i.e., `$r$-sets') are defined~\cite{requicha:80} as compact (bounded and closed) regular and semianalytic subsets of the $d$-space. A large number of representation schemes were defined in the past forty years, including the two main classes of (a) \emph{boundary representations} (`$B$-reps'), where the solid model is represented through a representation of its boundary elements, i.e.~faces, edges and vertices, and (b) \emph{decompositive/enumerative representations}, that are a decomposition of either the object or the embedding space, respectively, into a well-defined cellular complex. In particular, a boundary representation provides a cellular decomposition of the object's boundary into cells of dimension zero (vertices), one (edges), and two (faces). Medical imaging is a kind of enumerative representation of organs and tissues of interest, as subsets of 3D volume elements (voxels) from the 3D image.

>>>>

A \emph{representation scheme} for solid modeling is a mapping between a space of mathematical models and a space of symbolic representations, often generated by a formal grammar.
Solid pointsets (i.e., `$r$-sets') are defined~\cite{requicha:80} as compact (bounded and closed) regular and semianalytic subsets of the $d$-space. A large number of representation schemes were defined in the past forty years, including the two main classes of (a) \emph{boundary representations} (`$B$-reps'), where the solid model is represented through a representation of its boundary elements, i.e.~faces, edges and vertices, and (b) \emph{decompositive/enumerative representations}, that are a decomposition of either the object or the embedding space, respectively, into a well-defined cellular complex. In particular, a boundary representation provides a cellular decomposition of the object's boundary into cells of dimension zero (vertices), one (edges), and two (faces). Medical imaging can be classified as \emph{enumerative representation} of cellular decompositions of organs and tissues of interest, in particular, as subsets \emph{of 3D volume elements} (voxels) from the 3D image.

### 2019-06-11 src/implementation.tex
posted contents for implementation section

>>>>

The computer code is implemented in Julia language~\cite{} according to the workflow described below, whose stages are parallelized and/or optimized in various ways.

\paragraph{Workflow setup} The functions in this preliminary step include:
\begin{enumerate}

\item input of 3D medical image $\mathcal{I}$ dimensions $\ell_1, \ell_2, \ell_3$, such that: $\mathcal{I} = [\ell_1]\times[\ell_2]\times[\ell_3]$, where $[\ell_k] = (1,2,\ldots,\ell_k)$;

\item analysis of resources available in the computational environment, including operating system, type and number of compute nodes (processors, cores, GPUs), number of cores per node, RAM and caches amounts;

\item depending on the above, best decision for $\emph{size}$ of 3D image blocks $B$ (bricks); defaults to: $\emph{size}=64$; hence the number of bricks will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$. 
Hence default value: $n=256$, for images $\mathcal{I} = 512\times 512\times 256$;

\item computation of a Julia's sparse boundary matrix $[\partial_B]$, returning a value of type \texttt{SparseMatrixCSC\{Int8\}\{Int64\}}, where \texttt{Int8} and \texttt{Int64} are the types for values and indices, respectively, stored by Compressed Sparse Column (CSC) format; the average $[\partial_B]$ (for $\emph{size}$ = 64) is about

\item creation of a either a local or distributed channel to implement a producer/consumer model of parallel/distributed computation, depending on available resources;

\item distribution of matrix $[\partial_B]$, of size $45$ MB, to all available nodes/cores (workers) using the macro \texttt{@eveywhere}.

\end{enumerate}

The number of non-zeros within the sparse matrix $[\partial_{64^3}]$ is $\mathtt{nnz} = 4 792 266$, for a memory size of $9\times \mathtt{nnz}+8*262144 \simeq 45$ MB. The memory size of the sparse matrix is computed as $8+1$ bytes for non-zero element (i.e. 6 per row), plus 8 bytes per column.

\paragraph{Job enqueuing} Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that provide producer/consumer communication. Overall execution time can be improved if other tasks can be run while a task is being executed, or while waiting for an external service/function to complete. The single work items follows:
\begin{enumerate}

\item extraction of the block views from image arrays depending on 3 indices;

\item transformation of each job from global to local coordinates;

\item further transform of each forground voxel from Cartesian to linear coordinates, using the suitable Julia's library functions.

\item enqueing of the job (a sequence of integer positions for the non-zeros image elements aligned in a memory buffer).
\end{enumerate}

\paragraph{3-Chain encoding} 
Each block $B$ of the 3D image must by converted into the coordinate representation of a vector of the linear space $C_3$ of 3-chains. In coordinates local to $B$, once fixed an ordering from Cartesian to linear coords, this vector is represented by a binary array of length $64^3=262144$,  with a non-zero value (i.e.~$1$) for each foreground voxel in the interesting image portion in $B$. Therefore, the coded portion of the segment $\mathcal{S}$ inside $B$, denoted as $\mathcal{S}(B)$, results with a space occupancy of about $262$ KB if encoded as a full array (i.e.~including the zero values), When encoded as a sparse vector, its space occupancy will correspondingly decrease.

\begin{enumerate}

\item each job encoding produces either a full or sparse binary vector. With full arrays depending by one index, we get 262 KB per job;

\item special format for sparse CSC (Compressed Sparse Column) vectors can be used, since the \emph{value} data for non-zeros do not need storage. Hence only a single 1-array of \texttt{Int64} row positions (with total length equal to the number of non-zeros in the block, with $8\times\mathtt{nnz}$ KB storage) is needed;

\item prepare subsequnces of such data vectors (non-zero row indices in linear indices), in order to feed efficiently the available processor threads.
In case of presence of one/more GPUs, a smaller size of the block---and hence of the boundary matrix and the encoded 3-chain vectors---and then much higher vector numbers, are preferable for speed.

\end{enumerate}

\paragraph{SpMM Multiplication} 
According to the current literature~\cite{} it is more convenient to execute SpMV (sparse matrix-vector) multiplications than SpMSpV (sparse matrix-sparse vector) multiplications. Since we have 256 such jobs (one multiplication per block) to perform in the standard setting of the algorithm\footnote{
Size of block $64^3$; size of image $512^2\times 256$.
}, or even more, in case of either smaller blocks or image greater than the standard ones, this stage must be evidently parallelized and carefully tuned, possibly by using the available GPUs.
\begin{enumerate}

\item Various multiplication algorithms are being experimented, using several packages for sparse algebra and/or customized implementations;

\item the total speed of this stage will strongly depend on both the hardware available, on the granularity of blocks, and on the choice between dense/sparse storage of encoded 3-chains;

\item anyway, the compute elements or threads will be feed without solution of continuity in a \emph{dataflow} process. This parallel operation is, according to our preliminary experiments, the critical one of the whole workflow, in the sense that any $\Delta T$ (either positive or negative) in this stage will contribute to the total time $T$.

\end{enumerate}

\paragraph{2-Chain decoding} 
Each multiplication of $[\partial_B] : C_3 \to C_2$, times a 3-chain $\nu\in C_3$, produces a 2-chain  $\sigma\in C_2$, i.e.~the coordinate representation of the boundary vector $\sigma\in C_2$.  The inverse of the coding algorithm is executed in the present stage.  This process can also be partially superimposed in time with the previous one, depending on the size of the memory buffer used to feed the CPU cores or the GPUs. The elementary steps follow:

\begin{enumerate}

\item conversion from position of ones (or non-zeros) in the 2-chain to row linear indices;

\item conversion from linear indices to Cartesian indices in coordinates local to the $B$ block, using the appropriate library functions;

\item conversion from each Cartesian index value to a suitably oriented (i.e.~with proper attitude) geometry quadrilateral (or pair of triangles) in local coordinates.

\end{enumerate}

A Julia's vectorized pipeline seems the more appropriate implementation model for the job of each worker.

\paragraph{Assembling and artifact filtering} 
The results of the previous stages can be described a a collection of sets of geometric quads, each one encoded as an array of quadruples of integer indices, pointing to the linear array of grid vertices associated to the standard image block $B$.  In other words, \emph{all quads} of \textbf{each job} are now given in the \textbf{same} \emph{local coordinates}.  Other than to put each partial surface $\mathcal{S}(B) = (\texttt{V}_B, \sum_{\sigma\in B} \texttt{FV}_\sigma)$ in the global coordinate system of the image, the present stage must eliminate the redundant boundary features possibly generated at the edges of the partial surface $\mathcal{S}(B)$ within each block $B$ such that $B \cap \mathcal{I} \not= \emptyset$.
\begin{enumerate}

\item translate each array \texttt{FV}, of type \texttt{Lar.Cells}, by summing to each linear vertex index the linearized offset of the Cartesian coordinates $(n,m,p)(B)$ of the vertex with lowest coordinates of its block $B$.

\item remove both instances of double quads generated by \texttt{Lar} software at the block boundaries (see Figure~\ref{fig:blockboundary}). They are artifacts generated by the decomposition of the whole image into a number of blocks of tractable size.

\item 
a smart strategy of removal of such artifacts can be used, which does not require any sorting nor searching on the assembled array of quads. It will consist of arranging each block with all three dimensions increased by one, so that each 2-adjacent pair of blocks will covering each other for a full side of blocks of depth one. The details of this artifact filtering will be elucidated in Section~\ref{sec:covering}.

\end{enumerate}

\paragraph{Smoothing} 
The final smoothing of the generated surfaces cannot be performed block-wise, since this would introduce smoothing artifacts at the block boundaries. Anyway, the Taubin smoothing~\cite{} can be performed in parallel, since for each vertex in the final surface (except eventually the ones on the image $\mathcal{I}$ boundaries) it essentially consist in computing a new position as a proper average of its neighborhood vertices, i.e.~by applying a discrete Laplacian operator.  Some appropriate set workers may so be assigne the task of generating iteratively a new position for the vertices they take cure of. In particular, we have:
\begin{enumerate}

\item Job enqueuing, by writing sets of integers (global linear indices of vertices) in array buffers of appropriate type \texttt{Channel};

\item iterated vectorized computation of proper averages of closed vertices;

\item job dequeuing, by recovering finished tasks from a channel and assembling the results into the embedding function $\mathtt{V}: C_0 \to \E^3$ providing an array of type \texttt{Lar.Points} of \texttt{Float64 $\times$ 3}, with vertex coordinates by column.
\end{enumerate}
